---
title: "Josephs-Final"
author: "Moiya Josephs"
date: "2022-10-18"
output: pdf_document
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(curl)
library(psych)
```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

Does having a social life predict final grade? I would use t-test to answer this question, comparing the groups that have a social life vs those who do not.


### Cases 

**What are the cases, and how many are there?**

There are 649 cases that represent students and their achievements in secondary education in two Portuguese schools.

### Data collection 

**Describe the method of data collection.**

The data was collected using school reports and questionnaires.

### Type of study 

**What type of study is this (observational/experiment)?**

This is observational study.

### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

The data was collected by University of Minho and the dataset can be found here: 
Source: https://archive.ics.uci.edu/ml/datasets/Student+Performance


### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variable is the final grade of each student. It is quantitative.

### Independent Variable(s)

The independent variable is the variables that describe the students social life. To answer this questions the variables I classify as describing a students social life is their activities, romantic, family relationship, free time, going out and alcohol consumption. They are all qualitative.



### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}
student_mat_csv <- "https://raw.githubusercontent.com/moiyajosephs/Data606-Final/main/student-mat.csv"
student_mat <- read_delim(curl(student_mat_csv),delim = ";")
```



```{r}
summary(student_mat)
```


```{r}
ggplot(student_mat, aes(sex)) + geom_bar()
```

#### Activities

Students reply yes or no if they have extra curricular activities, indicated by the `activities` variables.

```{r}
table(student_mat$activities)
ggplot(student_mat, aes(activities)) + geom_bar()
```

#### Romantic

Students reply either yes or no if they are in a romantic relationship, indicated by the `romantic` variable.

```{r}

ggplot(student_mat, aes(romantic)) + geom_bar()
```

### Freetime 

Freetime after school rated from 1 very low to 5 very high.

```{r}
ggplot(student_mat, aes(freetime)) + geom_bar()

```

#### Family Relationship

`famrel` indicated the quality of family relationships on a scale from (numeric: from 1 - very bad to 5 - excellent)


```{r}
ggplot(student_mat, aes(famrel)) + geom_bar()

```


#### Going Out

`goout` indicates if students go out with friends rated from (numeric: from 1 - very low to 5 - very high)


```{r}
ggplot(student_mat, aes(goout)) + geom_bar()

```


#### Alcohol Consumption

Dalc indicated the workday alcohol consumption from very low to very high and Walc indicates weekend alcohol consumption, also from very low to very high.


```{r}
ggplot(student_mat, aes(Dalc)) + geom_bar()

```


```{r}
ggplot(student_mat, aes(Walc)) + geom_bar()

```

#### Grades

```{r}
ggplot(student_mat, aes(x=G3)) + geom_histogram()

```


#### Sample comparison 

For the final project I will compare the final grades with the other social factors. Here I compare with the student romantic social life.

```{r}
describe(student_mat$G3)
```


```{r}
describeBy(student_mat$G3, 
           group = student_mat$romantic, mat=TRUE)
```



# Example with null hypothesis 


To remove any relationship that might exist in the population(by permutation)

Calculate the original observed statistic. 

```{r}

# This would answer is there a difference in romantic relationships between sex and romantic relationships

first.one <- student_mat %>%
  select(sex,romantic)  %>%
  filter(romantic %in% c("yes","no"))

diff_orig <- student_mat %>% 
  group_by(sex) %>%
  summarize(prop_romance = mean(romantic == "yes")) %>%
  summarize(obs_diff_prop = diff(prop_romance))

diff_orig
```



# Sex with grades


```{r}
# Difference between gender and their grades

## change G3, to pass or fail
## if over a certain grade, pass, else fail. then find the difference in propostion of high grades
second.one <- student_mat %>%
  select(sex,romantic, G3)

diff_orig2 <- student_mat %>% 
  group_by(sex, romantic) %>%
  # summarize the proportion of people who passed
  summarize(prop_final = mean(G3>15)) %>%
  # summarize the difference in proportion of high grades
  summarize(obs_diff_prop = diff(prop_final))

diff_orig2


# gender and their grades and their romance

```

```{r}
library(infer)
```


```{r}
model <- lm(G3 ~ romantic, data = student_mat)
summary(model)
```


```{r}
ggplot(student_mat, aes(x=romantic, y = G3)) +
  geom_jitter() +
  geom_point(alpha = 0.2)
```

```{r}
ggplot(student_mat, aes(x=romantic, y = G3)) +
  geom_boxplot()
```

```{r}
ggplot(student_mat, aes(x= G3)) +
  geom_histogram(bins = 10) +
  facet_wrap(vars(romantic))
```

```{r}
summary_stats <- student_mat %>%
  group_by(romantic) %>%
  summarise(mean_by_group = mean(G3))
summary_stats
```


## Making predictions: 

```{r}
explanatory_data <- tibble(romantic = c("yes","no"))
explanatory_data
```

```{r}
explanatory_data %>%
  mutate(
    respnse_var = predict(model, explanatory_data)
  )
```

```{r}
predict(model,explanatory_data)
```


## Visualizing predictions
```{r} 
#Predict the grade for each student
predict(model)
```




```{r}
model <- lm(G3 ~ activities, data = student_mat)
summary(model)
```





```{r}
model <- lm(G3 ~ paid, data = student_mat)
summary(model)
```

