---
title: "Josephs-Final"
author: "Moiya Josephs"
date: "2022-10-18"
output: pdf_document
---

## Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(curl)
library(psych)
```

## Abstract


## Research question 

Does having a social life predict final grade? 


## Cases 

**What are the cases, and how many are there?**

There are 649 cases that represent students and their achievements in secondary education in two Portuguese schools.

## Data collection 

**Describe the method of data collection.**

The data was collected using school reports and questionnaires.

## Type of study 

**What type of study is this (observational/experiment)?**

This is observational study.

## Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

The data was collected by University of Minho and the dataset can be found here: 
Source: https://archive.ics.uci.edu/ml/datasets/Student+Performance


### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variable is the final grade of each student. It is quantitative.

### Independent Variable(s)

The independent variable is the variables that describe the students social life. To answer this questions the variables I classify as describing a students social life is their activities, romantic, family relationship, free time, going out and alcohol consumption. They are all qualitative.



## Relevant summary statistics 


```{r}
student_mat_csv <- "https://raw.githubusercontent.com/moiyajosephs/Data606-Final/main/student-mat.csv"
student_mat <- read_delim(curl(student_mat_csv),delim = ";")
```


```{r}
summary(student_mat)
```


```{r}
ggplot(student_mat, aes(sex)) + geom_bar()
```


## Romantic

Students reply either yes or no if they are in a romantic relationship, indicated by the `romantic` variable.

```{r}

ggplot(student_mat, aes(romantic)) + geom_bar()
```

## Freetime 

Freetime after school rated from 1 very low to 5 very high.

```{r}
ggplot(student_mat, aes(freetime)) + geom_bar()

```


## Grades



```{r}
describe(student_mat$G3)
```




## The Romance Model

```{r}
romantic.model <- lm(G3 ~ romantic, data = student_mat)
summary(romantic.model)
```


```{r}
ggplot(student_mat, aes(x=romantic, y = G3)) +
  geom_jitter() +
  geom_point(alpha = 0.2)
```

```{r}
ggplot(student_mat, aes(x=romantic, y = G3)) +
  geom_boxplot()
```

```{r}
ggplot(student_mat, aes(x= G3)) +
  geom_histogram(bins = 10) +
  facet_wrap(vars(romantic))
```

```{r}
summary_stats <- student_mat %>%
  group_by(romantic) %>%
  summarise(mean_by_group = mean(G3))
summary_stats
```


```{r}
describeBy(student_mat$G3, 
           group = student_mat$romantic, mat=TRUE)
```


### Making predictions: 

```{r}
explanatory_data <- tibble(romantic = c("yes","no"))
explanatory_data
```

```{r}
explanatory_data %>%
  mutate(
    response_var = predict(romantic.model, explanatory_data)
  )
```



### Visualizing predictions
```{r} 
#Predict the grade for each student
predict(romantic.model)
```



## The Activities Model
Students reply yes or no if they have extra curricular activities, indicated by the `activities` variables.

```{r}
ggplot(student_mat, aes(activities)) + geom_bar()
```

```{r}
activities.model <- lm(G3 ~ activities, data = student_mat)
summary(activities.model)
```



```{r}
ggplot(student_mat, aes(x=activities, y = G3)) +
  geom_boxplot()
```

```{r}
ggplot(student_mat, aes(x= G3)) +
  geom_histogram(bins = 10) +
  facet_wrap(vars(activities))
```

```{r}
summary_stats <- student_mat %>%
  group_by(activities) %>%
  summarise(mean_by_group = mean(G3))
summary_stats
```


### Making predictions: 

```{r}
explanatory_data <- tibble(activities = c("yes","no"))
explanatory_data
```

```{r}
explanatory_data %>%
  mutate(
    respnse_var = predict(activities.model, explanatory_data)
  )
```


## Freetime 

Freetime after school rated from 1 very low to 5 very high.

```{r}
ggplot(student_mat, aes(freetime)) + geom_bar()

```

```{r}
freetime.model <- lm(G3 ~ freetime, data = student_mat)
summary(freetime.model)
```
Not significant!



## Family Relationship

`famrel` indicated the quality of family relationships on a scale from (numeric: from 1 - very bad to 5 - excellent)


```{r}
ggplot(student_mat, aes(famrel)) + geom_bar()

```

```{r}
famrel.model <- lm(G3 ~ famrel, data = student_mat)
summary(famrel.model)
```

No significance

## Going Out

`goout` indicates if students go out with friends rated from (numeric: from 1 - very low to 5 - very high)


```{r}
ggplot(student_mat, aes(goout)) + geom_bar()

```


```{r}
goout.model <- lm(G3 ~ goout, data = student_mat)
summary(goout.model)

```

Significance!!



## Alcohol Consumption

Dalc indicated the workday alcohol consumption from very low to very high and Walc indicates weekend alcohol consumption, also from very low to very high.


```{r}
ggplot(student_mat, aes(Dalc)) + geom_bar()

```


```{r}
Dalc.model <- lm(G3 ~ Dalc, data = student_mat)
summary(Dalc.model)
```

Not significant


```{r}
ggplot(student_mat, aes(Walc)) + geom_bar()

```


```{r}

Walc.model <- lm(G3 ~ Walc, data = student_mat)
summary(Walc.model)
```


Not significant



## Academic Help



```{r}
model <- lm(G3 ~ paid, data = student_mat)
summary(model)
```






### Domestic Life


# Conclusion
Why is this analysis important?
Limitations of the analysis?

